A density ratio framework for evaluating the utility of synthetic data

### 2024-08-23 paper/paper.qmd
Update paper

---
title: "A density ratio framework for evaluating the utility of synthetic data"
format:
  arxiv-pdf:
    keep-tex: true
    include-in-header:
      - file: preamble.tex
  arxiv-html: default
number-sections: true
linenumbers: false
doublespacing: true
runninghead: "Density ratios for utility"
authorcols: true
link-citations: true
author:
  - name: Thom Benjamin Volker
    affiliations:
      - name: Utrecht University | Statistics Netherlands
        department: Methodology and Statistics | Methodology
        address: Padualaan 14
        city: Utrecht
        country: The Netherlands
        postal-code: "3584CH"
    orcid: 0000-0002-2408-7820
    email: t.b.volker@uu.nl
  - name: Peter-Paul de Wolf
    affiliations:
      - name: Statistics Netherlands
        department: Methodology
        address: PO Box 24500
        city: The Hague
        country: The Netherlands
        postal-code: "2490HA"
    email: pp.dewolf@cbs.nl
  - name: Erik-Jan van Kesteren
    affiliations:
      - name: Utrecht University
        department: Methodology and Statistics
        address: Padualaan 14
        city: Utrecht
        country: The Netherlands
        postal-code: "3584CH"
    email: e.vankesteren1@uu.nl
abstract: TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO 
keywords: |
  synthetic data, utility, density ratio, privacy, disclosure limitation
bibliography: bibliography.bib  
---

@@

@@

Global utility measures typically attempt to quantify the distributional similarity between the observed and synthetic data samples.
The intuition is that if two data sets have similar distributions, the data sets can be used for the same purposes, and analyses on the two data sets should yield similar results.
A common way to formalize distributional similarity is through the Kullback-Leibler (KL) divergence, as proposed in @karr_utility_2006.
The KL-divergence measures the relative entropy from the probability distribution of the observed data $\pobs(\bx)$ to the probability distribution of the synthetic data $\psyn(\bx)$ (with $\bx \in \mathbb{R}^{n \times d}$), and is defined as
$$
D_{\text{KL}}(\psyn || \pobs) = 
\int \psyn(\bx) \log \frac{\psyn(\bx)}{\pobs(\bx)} \text{d} \bx.
$${#eq-kl-div}
@karr_utility_2006 argue that the KL-divergence can be constructed from density estimators $\hat{p}_\text{obs}$ and $\hat{p}_\text{syn}$, after which the integral can be approximated using numerical quadrature.
Alternatively, if the observed and synthetic data are both (multivariate) normally distributed, the KL-divergence can be computed analytically.
Both implementations might be unfeasible: the combination of density estimation with numerical quadrature is cumbersome in high dimensional settings, and assuming multivariate normality might be too restrictive.
Yet, we show in later sections that the density ratio estimation framework gives rives to an alternative way of estimating the KL-divergence that overcomes these challenges.

An alternative way to evaluate whether two distributions are statistically indistinguishable is by evaluating whether a classification model can tell samples from the two distributions apart [see @kim_classification_2021, who formalize the connection between classification accuracy and two-sample testing].
In the context of synthetic data, this implies that if a classification model can distinguish observed from synthetic samples, the distributional similarity is low, and so is the global utility. 
If a classifier cannot distinguish between the observed and synthetic data, one would conclude that the global utility is high.
The propensity score mean-squared error ($pMSE$), introduced by @Woo_global_2009 and further developed in @snoke_utility_2018, formalizes this intuition. 
Let $I_i$ denote an indicator variable that equals $1$ if observation $i$ ($i \in 1, \dots, N$, $N = \nobs + \nsyn$) belongs to the synthetic data, and $0$ otherwise.
We then train a classifier that outputs the predicted probability of observation $i$ being a synthetic record $\hat{s}_i$ based on the observation's scores on the variables (this can be the set of all variables, but also a subset).
From these, we can calculate the utility statistic 
$$
pMSE = \frac{1}{N} \sum_{i=1}^N \Big(\hat{s}_i - \frac{\nsyn}{N}\Big)^2,
$$ {#eq-pMSE}
which ought to be smaller when the synthetic data is more like the observed data.
Crucially, the $pMSE$ depends on the classification model used and increases in the flexibility of the classification model, making it prone to overfitting and hard to interpret.
To combat these issues, @snoke_utility_2018 suggests to compare the $pMSE$-value with its expectation under the null distribution.
Provided that the classification model is a logistic regression model with $k$ parameters (including the intercept), @snoke_utility_2018 show that the expected $pMSE$ is given by
$$
\mathbb{E}[pMSE] = \Big(\frac{k-1}{N}\Big) \Big(\frac{\nobs}{N}\Big)^2
\Big(\frac{\nsyn}{N}\Big).
$$
For other classification models, the expectation can be approximated through a resampling procedure. 
Accordingly, the $pMSE$-ratio is given by
$$
pMSE\text{-ratio} = \frac{pMSE}{\mathbb{E}[pMSE]},
$$
with values smaller than $10$ deemed acceptable [@raab2021assessing\; although the authors remark that the smaller the better].
Apart from the $pMSE$, several other measures exist that can be constructed from the estimated propensity scores, such as the percentage of records correctly predicted [@raab2021assessing] or the Kolmogorov-Smirnov statistic [@Bowen_differentially_2021], both which are strongly correlated with the $pMSE$ [@raab2021assessing].

Due to its intuitive nature, multiple studies advice the use of the $pMSE$ as a promising technique to evaluate the quality of synthetic data [e.g., @raab2017guidelines; @raab2021assessing; @hu_advancing_2024].
Yet, it is not free of criticism. 
The usefulness of the $pMSE$ hinges on choosing a model that can capture the important intricacies of the observed data. 
@drechsler_utility_2022 illustrated that the utility score is highly dependent on the model used to estimate the propensity scores, and that clear improvements in synthesis models are not necessarily picked up in the $pMSE$.
Moreover, the $pMSE$ is prone to overfitting and users may find it difficult to select an appropriate model for estimating the propensity scores.
Hence, current methods to evaluate global utility are either difficult to estimate, especially in high-dimensional situations, or depend strongly on model specification.
Density ratio estimation might alleviate these issues.
In what follows, we explain how the density ratio can be estimated, and how it can be used to evaluate the utility of synthetic data.

@@

The density ratio estimation framework was originally developed in the machine learning community for the comparison of two probability distributions [for an overview, see @sugiyama_suzuki_kanamori_2012]. 
The framework has been shown to be applicable to prediction [@sugiyama_conditional_2010; @sugiyama_classification_2010], outlier detection [@shohei_dre_outlier_2008], change-point detection in time-series [@liu_change_2013], importance weighting under domain adaptation [i.e., sample selection bias\; @kanamori_ulsif_2009], and two-sample homogeneity tests [@sugiyama_lstst_2011]. 
The general idea of density ratio estimation is depicted in @fig-dr-plot, and boils down to comparing two distributions by modelling the density ratio $r(\bx)$ between the probability distributions of the numerator samples, taken from the synthetic data distribution, $\psyn(\bx)$, and the denominator samples, taken from the observed data distribution, $\pobs(\bx)$, such that
$$
r(\bx) = \frac{\psyn(\bx)}{\pobs(\bx)}.
$$ {#eq-dr}

This specification has the intuitive interpretation that in locations where the density ratio takes large values, too many synthetic observations will be generated in that region and at the locations where the density ratio is small, there will be too few synthetic observations, both relative to the observed data. 
An intuitive approach to estimating $r(\bx)$ from samples of $\pobs(\bx)$ and $\psyn(\bx)$ would be to estimate the observed and synthetic data density separately, for example using kernel density estimation [e.g., see @Scott1992 for an overview], and subsequently compute the ratio of these estimated densities.
However, density estimation is one of the hardest tasks in statistical learning, unavoidably leading to estimation errors for both densities, especially in high dimensions.
When subsequently taking the ratio of the estimated densities, estimation errors tend to be magnified. 
Direct density ratio estimation avoids this issue by specifying and estimating a model directly for the ratio without first estimating the separate densities.
Extensive simulations on a wide variety of tasks showed that this approach typically outperforms density ratio estimation through naive kernel density estimation, especially when the dimensionality of the data increases [e.g., @Kanamori2012; @shohei_dre_outlier_2008; @kanamori_ulsif_2009].

```{r}
#| label: fig-dr-plot
#| echo: false
#| warning: false
#| message: false
#| fig-dpi: 500
#| fig-cap: "Example of the true and estimated density ratio of two normal distributions with different means and variances (i.e., $\\psyn(\\bx) = N(0,1)$ and $\\pobs(\\bx) = N(1,2)$). The function $r(\\bx) = \\psyn(\\bx)/\\pobs(\\bx)$ denotes the true density ratio, the function $\\hat{r}(\\bx)$ denotes an estimate of the density ratio based on $\\nsyn = \\nobs = 200$ samples from each distribution obtained with unconstrained Least-Squares Importance Fitting (uLSIF). Note that the density ratio is itself not a proper density."
#| out-width: 100%
#| fig-height: 1.8
#| fig-pos: t
#| dev: cairo_pdf

@@

ggplot() +
  stat_function(aes(col = "A"), 
                fun = dnorm, args = list(mean = 0, sd = 1),
                size = 0.5) +
  stat_function(aes(col = "B"),
                fun = dnorm, args = list(mean = 1, sd = sqrt(2)),
                size = 0.5) +
  scale_color_brewer(labels = c("A" = expression(italic(p)[syn](x)),
                                "B" = expression(italic(p)[obs](x))),
                     palette = "Set2",
                     name = "") +
  xlim(-4, 6) +
  ylim(0, 0.5) +
  ylab("Density") +
  theme_minimal() +
  theme(legend.position = "bottom",
        text = element_text(family = "LM Roman 10"),
        legend.background = element_blank()) +
ggplot() +
  stat_function(aes(col = "C"),
                fun = ~ dnorm(.x, 0, 1) / dnorm(.x, 1, sqrt(2)),
                size = 0.5) +
  stat_function(aes(col = "D"),
                fun = ~ predict(rhat, newdata = .x),
                size = 0.5) +
  scale_color_manual(labels = c("C" = expression(italic(r(x))),
                                "D" = expression(italic(hat(r)(x)))),
                     values = c("black", "steelblue3"),
                     name = "") +
  xlim(-4, 6) +
  ylim(0, 2.5) +
  theme_minimal() +
  ylab("Density ratio") +
  theme(legend.position = "bottom",
        text = element_text(family = "LM Roman 10"))
```

@@

Over the past years, several methods for direct density ratio estimation have been developed.
A large class of these methods attempt to directly minimize the error between the true density ratio $r(\bx)$ and its estimate $\hat{r}(\bx)$.
Following this approach, we define a loss function $\mathcal{L}(r(\bx), \hat{r}(\bx))$ that measures the discrepancy between the true and estimated density ratio. 
To give an example, consider the following loss based on the squared error
$$
\begin{aligned}
\mathcal{L}_S(r(\bx), \hat{r}(\bx)) &= 
\frac{1}{2} \int (\hat{r}(\bx) - r(\bx))^2 \pobs(\bx) \text{d}\bx \\
&= \frac{1}{2} \int \hat{r}(\bx)^2 \pobs(\bx) \text{d}\bx - 
\int \hat{r}(\bx) r(\bx) \pobs(\bx) \text{d}\bx + 
\frac{1}{2} \int r(\bx)^2 \pobs(\bx) \text{d}\bx.
\end{aligned}
$$ {#eq-squared-error}
The second term can be rewritten, because the denominator in $r(\bx)$ cancels with $\pobs(\bx)$, while the third term is a constant with respect to the parameters in the density ratio model and can thus be ignored. 
Hence, we are left with the following loss function to minimize
$$
\mathcal{L}'_S (r(\bx), \hat{r}(\bx)) = \frac{1}{2} \int \hat{r}(\bx)^2 \pobs(\bx) \text{d}\bx - \int \hat{r}(\bx) \psyn(\bx) \text{d}\bx,
$$ {#eq-squared-error-loss}
which is termed least squares importance fitting (LSIF) by @kanamori_ulsif_2009.

This approach can be straightforwardly generalized to a wide class of loss functions that fall under the family of Bregman divergences [see @sugiyama_bregman_2012; @mohamed2017learning]. 
Then, a general class of losses is encompassed by the expression
$$
\mathcal{L}_f (r(\bx), \hat{r}(\bx)) =
\int \Big(
f(r(\bx)) - f(\hat{r}(\bx)) - f'(\hat{r}(\bx))(r(\bx) - \hat{r}(\bx))
\Big) \pobs(\bx) \text{d}\bx,
$$ {#eq-bregman-loss}
where $f$ is a differentiable and strictly convex function with derivative $f'$.
Then, ignoring the terms independent of $\hat{r}(\bx)$ and noting that $r(\bx)\pobs(\bx) = \psyn(\bx)$, we obtain the following objective
$$
\mathcal{L}'_f (r(\bx), \hat{r}(\bx)) = \int 
\Big( 
f'(\hat{r}(\bx)) \hat{r}(\bx) - f(\hat{r}(\bx))
\Big)  \pobs(\bx) \text{d}\bx
- \int f'(\hat{r}(\bx)) \psyn(\bx)
\text{d}\bx.
$$ {#eq-bregman-objective}
Minimizing this loss for different functions $f$ yields different estimators for the density ratio, that focus on different regions of the density ratio.^[It is easy to see that using $f(x) = \frac{1}{2}(x-1)^2$ turns the Bregman divergence (@eq-bregman-objective) into the squared error (@eq-squared-error-loss).]
Specifically, some estimators place more emphasis on accurately modelling the regions in which the true density ratio is large, but less emphasis on accurately estimating the smaller density ratios, and vice versa [see @sugiyama_bregman_2012; @menon2016dreloss]. 
Importantly, @sugiyama_bregman_2012 show that the Bregman divergence minimization approach to density ratio estimation is equivalent to estimating the $f$-divergence [@ali_silvey_divergence_1966] between the synthetic and observed data distributions.
Hence, when estimating the density ratio, one implicitly estimates the divergence between the synthetic and observed data distributions.

After defining a loss function, we need a model for the density ratio function $\hat{r}(\bx)$. 
There are many possibilities to specify this model, but a common choice is to use a linear model for the density ratio [e.g., @huang_kmm_2006; @kanamori_ulsif_2009; @izbicki_dre_2014; @gruber2024overcoming]. 
That is, we define the density ratio model as
$$
\hat{r}(\bx) = \boldsymbol{\varphi}(\bx) \boldsymbol{\theta},
$$ {#eq-dr-model}
where $\boldsymbol{\varphi}(\bx)$ is a basis function vector, that transforms the data from a $p$-dimensional to a $b$-dimensional space, and $\boldsymbol{\theta}$ is a $b$-dimensional parameter vector.
Although the model is linear in its parameters, the density ratio itself is typically a non-linear function of the data due to the basis functions.
The parameter vector $\boldsymbol{\theta}$ is estimated to minimize the discrepancy with the true density ratio using the loss function $\mathcal{L}(r(\bx), \hat{r}(\bx))$ in @eq-bregman-objective.

Also for the basis functions, several specifications are possible, ranging from an identity function [@qin_inferences_1998] to normalizing flows [@choi_featurized_2021] or neural networks [@tiao2018dre; @uehara2016generative].
However, a commonly used basis function in the density ratio literature is the Gaussian kernel function [e.g., @huang_kmm_2006; @sugiyama_kliep_2007; @kanamori_ulsif_2009; @liu_change_2013; @gruber2024overcoming], which conveniently allows to model the non-linear density ratio using models that are linear in their parameters.
The Gaussian kernel function is defined as
$$
\boldsymbol{\varphi}(\bx) = \mathcal{K}(\bx_i, \boldsymbol{c}_j) = 
\exp\left(
- \frac{\lVert\bx - \boldsymbol{c}_j \rVert^2}{2 \sigma^2}
\right),
$$
where $\boldsymbol{c}_j$ ($j \in 1, \dots, J$) denotes the centers of the Gaussian kernel functions, and $\sigma$ controls the kernel width [i.e., it defines over which distance differences between the observations and the centers are considered relevant\; for more information on kernel functions, see, e.g., @murphy_pmlintro_2022].
The appropriate width of the kernel can be determined using cross-validation. 
The centers are typically sampled from the data (in our case, the observed and synthetic data can both be used, but it is also possible to solely use the synthetic samples).

>>>>

---
title: "A density ratio framework for evaluating the utility of synthetic data"
format:
  arxiv-pdf:
    keep-tex: true
    include-in-header:
      - file: preamble.tex
    header-includes:
      - \usepackage{float}
      - \floatplacement{table}{t}
  arxiv-html: default
number-sections: true
linenumbers: false
doublespacing: true
runninghead: "Density ratios for utility"
authorcols: true
link-citations: true
author:
  - name: Thom Benjamin Volker
    affiliations:
      - name: Utrecht University | Statistics Netherlands
        department: Methodology and Statistics | Methodology
        address: Padualaan 14
        city: Utrecht
        country: The Netherlands
        postal-code: "3584CH"
    orcid: 0000-0002-2408-7820
    email: t.b.volker@uu.nl
  - name: Peter-Paul de Wolf
    affiliations:
      - name: Statistics Netherlands
        department: Methodology
        address: PO Box 24500
        city: The Hague
        country: The Netherlands
        postal-code: "2490HA"
    email: pp.dewolf@cbs.nl
  - name: Erik-Jan van Kesteren
    affiliations:
      - name: Utrecht University
        department: Methodology and Statistics
        address: Padualaan 14
        city: Utrecht
        country: The Netherlands
        postal-code: "3584CH"
    email: e.vankesteren1@uu.nl
abstract: | 
  Synthetic data generation is a promising technique to facilitate the use of sensitive data while mitigating the risk of privacy breaches. However, for synthetic data to be useful in downstream analysis tasks, it needs to be of sufficient quality. Various methods have been proposed to measure the utility of synthetic data, but their results are often incomplete or even misleading. In this paper, we propose using density ratio estimation to improve quality evaluation for synthetic data, and thereby the quality of synthesized datasets.
  We show how this framework relates to and builds on existing measures, yielding global and local utility measures that are informative and easy to interpret. We develop an estimator which requires little to no manual tuning due to automatic selection of a nonparametric density ratio model. Through simulations, we find that density ratio estimation yields more accurate estimates of global utility than established procedures. A real-world data application demonstrates how the density ratio can guide refinements of synthesis models and can be used to improve downstream analyses. We conclude that density ratio estimation is a valuable tool in synthetic data generation workflows and provide these methods in the accessible open source R-package densityratio.
keywords: |
  synthetic data, utility, density ratio, privacy, disclosure limitation
bibliography: bibliography.bib
params:
  runsims: false
---

@@

```{r}
#| label: setup
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())

chunk_hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- chunk_hook(x, options)
  paste0("\\linespread{1}\n", x, "\n\n\\linespread{2}")
})
```

@@

Global utility measures typically attempt to quantify the distributional similarity between the observed and synthetic data samples.
The intuition is that two data sets are realizations from the same underlying distribution, the data sets can be used for the same purposes, and analyses on the two data sets should yield similar results. 
One way to evaluate distributional similarity is by assessing whether a classification model can tell samples from the two distributions apart [see @kim_classification_2021, who formalize the connection between classification accuracy and two-sample testing].
Hence, if a classification model can distinguish between the observed and synthetic data with high accuracy, the distributional similarity is low, and so is the global utility. 
If a classifier cannot distinguish between the observed and synthetic data, one would conclude that the global utility is high.
The propensity score mean-squared error ($pMSE$), introduced by @Woo_global_2009 and further developed in @snoke_utility_2018, formalizes this intuition.
Let $I_i$ denote an indicator variable that equals $1$ if observation $i$ ($i \in 1, \dots, N$, $N = \nobs + \nsyn$) belongs to the synthetic data, and $0$ otherwise.
We then train a classifier that outputs the predicted probability of observation $i$ being a synthetic record $\hat{\pi}_i$ based on the observation's scores on the variables (this can be the set of all variables, but also a subset).
From these, we can calculate the utility statistic 
$$
pMSE = \frac{1}{N} \sum_{i=1}^N \Big(\hat{\pi}_i - \frac{\nsyn}{N}\Big)^2,
$$ {#eq-pMSE}
which ought to be smaller when the synthetic data is more like the observed data.
Crucially, the $pMSE$ depends on the classification model used and increases in the flexibility of the classification model, making it prone to overfitting and hard to interpret.
To combat these issues, @snoke_utility_2018 suggests to compare the $pMSE$-value with its expected value under the null hypothesis that the observed and synthetic data are indistinguishable.
Provided that the classification model is a logistic regression model with $k$ parameters (including the intercept), @snoke_utility_2018 show that the expected $pMSE$ is given by
$$
\mathbb{E}[pMSE] = \Big(\frac{k-1}{N}\Big) \Big(\frac{\nobs}{N}\Big)^2
\Big(\frac{\nsyn}{N}\Big).
$$
For other classification models, the expectation can be approximated through a resampling procedure. 
Accordingly, the $pMSE$-ratio is given by
$$
pMSE\text{-ratio} = \frac{pMSE}{\mathbb{E}[pMSE]}.
$$
Apart from the $pMSE$, several other measures can be constructed from the estimated propensity scores, such as the percentage of records correctly predicted [@raab2021assessing] or the Kolmogorov-Smirnov statistic [@Bowen_differentially_2021], both which are strongly correlated with the $pMSE$ [@raab2021assessing].

Due to its intuitive nature, multiple studies advice the use of the $pMSE$ as a promising technique to evaluate the quality of synthetic data [e.g., @raab2017guidelines; @raab2021assessing; @hu_advancing_2024].
Yet, it is not free of criticism. 
The usefulness of the $pMSE$ hinges on choosing a model that can capture the important intricacies of the observed data. 
@drechsler_utility_2022 illustrated that the utility score is highly dependent on the model used to estimate the propensity scores, and that clear improvements in synthesis models are not necessarily picked up in the $pMSE$.
Moreover, selecting an appropriate propensity score model may be difficult, as the challenges associated with model selection, such as the bias-variance trade-off, typically apply.

Another way to formalize distributional similarity is through the Kullback-Leibler (KL) divergence, as proposed in @karr_utility_2006.
The KL-divergence measures the relative entropy from the probability distribution of the observed data $\pobs(\bx)$ to the probability distribution of the synthetic data $\psyn(\bx)$ (with $\bx \in \mathbb{R}^{d}$), and is defined as^[In contrast to @karr_utility_2006, we define the KL-divergence as the relative entropy of $\pobs$ with respect to $\psyn$, such that it is consistent with our later formulations.]
$$
D_{\text{KL}}(\pobs || \psyn) = 
\int \pobs(\bx) \log \frac{\pobs(\bx)}{\psyn(\bx)} \text{d} \bx.
$${#eq-kl-div}
@karr_utility_2006 argue that the KL-divergence can be computed by approximating the integral using density estimators $\hat{p}_\text{obs}$ and $\hat{p}_\text{syn}$. 
An implementation of this approach for divergence estimation has been described in @wang_divergence_2009, where the densities are estimated using nearest-neighbor density estimation. 
Another approach posed by @karr_utility_2006 is to assume multivariate normality for both $\psyn$ and $\pobs$, and calculate the KL-divergence in closed form from the sample means and covariance matrices. 
Although all these approaches are easy to implement, they have their limitations. 
Density estimation is challenging in high-dimensional settings, and assuming multivariate normality might be too restrictive. 
We show in later sections that the KL-divergence naturally fits into the density ratio estimation framework.

@@

The density ratio estimation framework was originally developed in the machine learning community for the comparison of two probability distributions [for an overview, see @sugiyama_suzuki_kanamori_2012]. 
The framework has been shown to be applicable to prediction [@sugiyama_conditional_2010; @sugiyama_classification_2010], outlier detection [@shohei_dre_outlier_2008], change-point detection in time-series [@liu_change_2013], importance weighting under domain adaptation [i.e., sample selection bias\; @kanamori_ulsif_2009], and two-sample homogeneity tests [@sugiyama_lstst_2011]. 
The general idea of density ratio estimation is depicted in @fig-dr-plot, and boils down to comparing two distributions by modelling the density ratio $r(\bx)$.
In our case, we define $r(\bx)$ as the ratio between the probability distributions of the numerator samples, taken from the observed data distribution, $\pobs(\bx)$, and the denominator samples, taken from the synthetic data distribution, $\psyn(\bx)$, such that
$$
r(\bx) = \frac{\pobs(\bx)}{\psyn(\bx)}.
$$ {#eq-dr}

This specification has the intuitive interpretation that in locations where the density ratio takes values larger than 1, too few synthetic observations are generated in that region.
Conversely, in locations where the density ratio takes values smaller than one, too many synthetic observations are generated in that region (see @fig-dr-plot).
We put $\pobs(\bx)$ in the numerator such that it plays the role of the actual distribution, while $\psyn(\bx)$ serves as the approximating distribution (we briefly come back to this in the next section).
A straightforward approach to estimating $r(\bx)$ from samples of $\pobs(\bx)$ and $\psyn(\bx)$ would be to estimate the observed and synthetic data density separately, for example using kernel density estimation [e.g., see @Scott1992 for an overview], and subsequently compute the ratio of these estimated densities.
However, density estimation is one of the hardest tasks in statistical learning, unavoidably leading to estimation errors for both densities, especially in high dimensions [@sugiyama_suzuki_kanamori_2012].
When subsequently taking the ratio of the estimated densities, estimation errors tend to be magnified. 
Direct density ratio estimation avoids this issue by specifying and estimating a model directly for the ratio without first estimating the separate densities.
Extensive simulations on a wide variety of tasks showed that this approach typically outperforms density ratio estimation through kernel density estimation for each distribution separately, especially when the dimensionality of the data increases [e.g., @Kanamori2012; @shohei_dre_outlier_2008; @kanamori_ulsif_2009].

```{r}
#| label: fig-dr-plot
#| echo: false
#| warning: false
#| message: false
#| fig-dpi: 500
#| fig-cap: "Example of the true and estimated density ratio of two normal distributions with different means and variances (i.e., $\\psyn(\\bx) = N(0,1)$ and $\\pobs(\\bx) = N(1,2)$). The function $r(\\bx) = \\pobs(\\bx)/\\psyn(\\bx)$ denotes the true density ratio, the function $\\hat{r}(\\bx)$ denotes an estimate of the density ratio based on $\\nsyn = \\nobs = 200$ samples from each distribution obtained with unconstrained Least-Squares Importance Fitting (uLSIF). Note that the density ratio is itself not a proper density."
#| out-width: 100%
#| fig-height: 1.8
#| fig-pos: t
#| dev: cairo_pdf

@@

ggplot() +
  stat_function(aes(col = "A", linetype = "A"), 
                fun = dnorm, args = list(mean = 0, sd = 1),
                size = 0.5) +
  stat_function(aes(col = "B", linetype = "B"),
                fun = dnorm, args = list(mean = 1, sd = sqrt(2)),
                size = 0.5) +
  scale_color_brewer(labels = c("A" = expression(italic(p)[obs](x)),
                                "B" = expression(italic(p)[syn](x))),
                     palette = "Set2",
                     name = "") +
  scale_linetype_manual(labels = c("A" = expression(italic(p)[obs](x)),
                                  "B" = expression(italic(p)[syn](x))),
                        values = c("solid", "dashed"),
                        name = "") +
  xlim(-4, 6) +
  ylim(0, 0.5) +
  ylab("Density") +
  theme_minimal() +
  theme(legend.position = "bottom",
        text = element_text(family = "LM Roman 10", size = 9),
        legend.background = element_blank()) +
ggplot() +
  geom_hline(mapping = aes(yintercept = 1), alpha = 0.3, size = 0.3) +
  stat_function(aes(col = "C", linetype = "C"),
                fun = ~ dnorm(.x, 0, 1) / dnorm(.x, 1, sqrt(2)),
                size = 0.5) +
  stat_function(aes(col = "D", linetype = "D"),
                fun = ~ predict(rhat, newdata = .x),
                size = 0.5) +
  scale_color_manual(labels = c("C" = expression(italic(r(x))),
                                "D" = expression(italic(hat(r)(x)))),
                     values = c("black", "steelblue3"),
                     name = "") +
  scale_linetype_manual(labels = c("C" = expression(italic(r(x))),
                                   "D" = expression(italic(hat(r)(x)))),
                        values = c("solid", "dashed"),
                        name = "") +  
  xlim(-4, 6) +
  ylim(0, 2.5) +
  theme_minimal() +
  ylab("Density ratio") +
  theme(legend.position = "bottom",
        text = element_text(family = "LM Roman 10", size = 9)) &
  theme(plot.margin = unit(c(0.2, 0.2, 0, 0.2), "cm"),
        legend.margin = unit(c(0.1, 0.1, 0.1, 0.1), "cm"))
```

@@

Over the past years, several methods for direct density ratio estimation have been developed.
A large class of these methods attempt to directly minimize the error between the true density ratio $r(\bx)$ and a density ratio model $\hat{r}(\bx)$.
Following this approach, we define a loss function $\mathcal{L}(r(\bx), \hat{r}(\bx))$ that measures the discrepancy between the true and estimated density ratio. 
To give an example, consider the following loss based on the squared error
$$
\begin{aligned}
\mathcal{L}^*_S(r(\bx), \hat{r}(\bx)) &= 
\frac{1}{2} \int (\hat{r}(\bx) - r(\bx))^2 \psyn(\bx) \text{d}\bx \\
&= \frac{1}{2} \int \hat{r}(\bx)^2 \psyn(\bx) \text{d}\bx - 
\int \hat{r}(\bx) r(\bx) \psyn(\bx) \text{d}\bx + 
\frac{1}{2} \int r(\bx)^2 \psyn(\bx) \text{d}\bx.
\end{aligned}
$$ {#eq-squared-error}
The second term can be rewritten, because the denominator in $r(\bx)$ cancels with $\psyn(\bx)$, while the third term is a constant with respect to the parameters in the density ratio model and can thus be ignored. 
Hence, we are left with the following loss function to minimize
$$
\mathcal{L}_S (r(\bx), \hat{r}(\bx)) = \frac{1}{2} \int \hat{r}(\bx)^2 \psyn(\bx) \text{d}\bx - \int \hat{r}(\bx) \pobs(\bx) \text{d}\bx,
$$ {#eq-squared-error-loss}
which is termed least squares importance fitting (LSIF) by @kanamori_ulsif_2009.

This approach can be straightforwardly generalized to a wide class of loss functions that fall under the family of Bregman divergences [see @sugiyama_bregman_2012; @mohamed2017learning]. 
Then, a general class of losses is encompassed by the expression
$$
\mathcal{L}^*_f (r(\bx), \hat{r}(\bx)) =
\int \Big(
f(r(\bx)) - f(\hat{r}(\bx)) - f'(\hat{r}(\bx))(r(\bx) - \hat{r}(\bx))
\Big) \psyn(\bx) \text{d}\bx,
$$ {#eq-bregman-loss}
where $f$ is a differentiable and strictly convex function with derivative $f'$.
Then, ignoring the terms independent of $\hat{r}(\bx)$ and noting that $r(\bx)\psyn(\bx) = \pobs(\bx)$, we obtain the following objective
$$
\mathcal{L}_f (r(\bx), \hat{r}(\bx)) = \int 
\Big( 
f'(\hat{r}(\bx)) \hat{r}(\bx) - f(\hat{r}(\bx))
\Big)  \psyn(\bx) \text{d}\bx
- \int f'(\hat{r}(\bx)) \pobs(\bx)
\text{d}\bx.
$$ {#eq-bregman-objective}
Minimizing this loss over $\hat{r}(\bx)$ for different functions $f$ yields different estimators for the density ratio, that focus on different regions of the density ratio.^[It is easy to see that using $f(x) = \frac{1}{2}(x-1)^2$ turns the Bregman divergence (@eq-bregman-loss) into the squared error (@eq-squared-error).]
Specifically, some estimators place more emphasis on accurately modelling the regions in which the true density ratio is large, but less emphasis on accurately estimating the density ratio in regions where it takes small values, and vice versa [see @sugiyama_bregman_2012; @menon2016dreloss]. 
Importantly, @sugiyama_bregman_2012 show that the Bregman divergence minimization approach to density ratio estimation is equivalent to estimating an $f$-divergence [@ali_silvey_divergence_1966] of the observed data distribution from the synthetic data distribution.
Hence, when estimating the density ratio, one implicitly estimates the divergence between the observed and synthetic data distributions.

After defining a loss function, we need a model for the density ratio function $\hat{r}(\bx)$. 
There are many possibilities to specify this model, but a common choice is to use a linear model for the density ratio [e.g., @huang_kmm_2006; @kanamori_ulsif_2009; @izbicki_dre_2014; @gruber2024overcoming]. 
That is, we define the density ratio model as
$$
\hat{r}(\bx) = \boldsymbol{\varphi}(\bx) \boldsymbol{\theta},
$$ {#eq-dr-model}
where $\boldsymbol{\varphi}(\bx)$ is a basis function vector (optionally including an intercept term), that transforms the data from a $p$-dimensional to a $b$-dimensional space, and $\boldsymbol{\theta}$ is a $b$-dimensional parameter vector.
Although the model is linear in its parameters, the density ratio itself is typically a non-linear function of the data due to the basis functions.
The parameter vector $\boldsymbol{\theta}$ is estimated to minimize the discrepancy with the true density ratio using the loss function $\mathcal{L}(r(\bx), \hat{r}(\bx))$ in @eq-bregman-objective.
Potentially, the loss function can be extended with a regularization term (e.g., a $L_1$ or $L_2$ norm on the parameters of the model), to decrease the flexibility of the density ratio.
Parameter estimation can be carried out using standard optimization techniques (see Appendix [-@sec-app-A] for a brief demonstration in `R`).

Also for the basis functions, several specifications are possible, ranging from an identity function [@qin_inferences_1998] to normalizing flows [@choi_featurized_2021] or neural networks [@tiao2018dre; @uehara2016generative].
A commonly used basis function in the density ratio literature is the Gaussian kernel function [e.g., @huang_kmm_2006; @sugiyama_kliep_2007; @kanamori_ulsif_2009; @liu_change_2013; @gruber2024overcoming].
The Gaussian kernel function is defined as
$$
\boldsymbol{\varphi}(\bx) = \mathcal{K}(\bx, \boldsymbol{c}_j) = 
\exp\left(
- \frac{\lVert\bx - \boldsymbol{c}_j \rVert^2}{2 \sigma^2}
\right),
$$
where $\boldsymbol{c}_j$ ($j \in 1, \dots, b$) denotes the centers of the Gaussian kernel functions, and $\sigma$ controls the kernel width [i.e., it defines over which distance differences between the observations and the centers are considered relevant\; for more information on kernel functions, see, e.g., @murphy_pmlintro_2022].
The appropriate width of the kernel and the optimal value of the regularization parameter can be determined using cross-validation. 
The centers are typically sampled from the data (in our case, both the observed and synthetic data).
